### 2017-07-04 reasoning.tex
Edits.

\begin{figure}
\centering
\begin{tabular}{l|l}
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  atomic {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{X = k! \conj Y = X\}
\end{decoration}$
\end{txnimpcode}
&
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  transaction {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v2 := v2 * v1;
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{Y = X\}
\end{decoration}$
\end{txnimpcode}
\end{tabular}
  \caption{Factorial program written as an atomic block and a weakly
  isolated transaction (shared variables have uppercase names).
  Effect of multiple iterations of the \C{atomic} loop can be
  summarized as $k!$, while same cannot be done with the
  \C{transaction} loop, where weak isolation allows reads in different
  iterations to return different values for $X$.  Nonetheless,
  post-state satisfies $X=Y$ in both cases due to atomicity of
  writes.}
\label{fig:atomic-vs-transaction}
\end{figure}

We now describe a proof system that lets us prove the correctness of a
\txnimp program $c$ w.r.t its high-level invariant $I$, on a machine
that satisfies the isolation specifications ($\I$) of its
transactions\footnote{Note the difference between $I$ and $\I$. The
former constitute \emph{proof} \emph{obligations} for the programmer,
whereas the latter describes a transaction's \emph{assumptions} about
operational characteristics of the underlying machine.}.  Our proof
system is essentially an adaptation of the rely-guarantee reasoning
framework to the setting of weakly isolated database transactions. It
deals with the additional complexity introduced by this setting, while
also taking advantage of the opportunities it provides to simplify the
reasoning. The primary challenge is the weak isolation; how do we
relate a transaction's isolation specification ($\I$) to its rely
relation ($R$) that describes the transaction's environment, so that
the interference is considered only insofar as the isolation level
allows it? Our proof system shows how to compute a rely relation ($R$)
modulo the isolation specification ($\I$) of a transaction. Another
characteristic of the transaction setting is atomicity; a
transaction's writes are not visible to its peers until it commits. In
the context of rely-guarantee, this means that the transactions
guarantee ($G$) should capture the aggregate effect of a transaction,
and not its individual writes.  While shared memory \C{atomic} blocks
also have the same characteristic, the fact that a transactions are
weakly isolated introduces complexity.  Unlike an \C{atomic} block,
the effect of a transaction is \emph{not} a sequential composition of
the effects of its statements because each statement witnesses a
potentially different version of the state. This is illustrated in
Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
a shared variable (\C{X}) can be composed to describe the effect of an
\C{atomic} block, while same cannot be done in a transaction. Thus,
weak isolation needs to be taken into account when defining and
verifying guarantees, and our proof system shows how.

>>>>

% \begin{figure}
% \centering
% \begin{tabular}{l|l}
% \begin{txnimpcode}
% $\begin{decoration}
%   \{X = k\}
% \end{decoration}$
%   atomic {
%     v1 := 0; v2 := 1;
%     while(v1<X) {
%       v2 := v2 * (X-v1);
%       v1 := v1+1;
%     }
%     X := v2; Y := v2;
%   }
% $\begin{decoration}
%   \{X = k! \conj Y = X\}
% \end{decoration}$
% \end{txnimpcode}
% &
% \begin{txnimpcode}
% $\begin{decoration}
%   \{X = k\}
% \end{decoration}$
%   transaction {
%     v1 := 0; v2 := 1;
%     while(v1<X) {
%       v2 := v2 * (X-v1);
%       v2 := v2 * v1;
%       v1 := v1+1;
%     }
%     X := v2; Y := v2;
%   }
% $\begin{decoration}
%   \{Y = X\}
% \end{decoration}$
% \end{txnimpcode}
% \end{tabular}
%   \caption{Factorial program written as an atomic block and a weakly
%   isolated transaction (shared variables have uppercase names).
%   Effect of multiple iterations of the \C{atomic} loop can be
%   summarized as $k!$, while same cannot be done with the
%   \C{transaction} loop, where weak isolation allows reads in different
%   iterations to return different values for $X$.  Nonetheless,
%   post-state satisfies $X=Y$ in both cases due to atomicity of
%   writes.}
% \label{fig:atomic-vs-transaction}
% \end{figure}

We now describe a proof system that lets us prove the correctness of a
\txnimp program $c$ w.r.t its high-level invariant $I$, on a machine
that satisfies the isolation specifications ($\I$) of its
transactions\footnote{Note the difference between $I$ and $\I$. The
former constitute \emph{proof} \emph{obligations} for the programmer,
whereas the latter describes a transaction's \emph{assumptions} about
operational characteristics of the underlying machine.}.  Our proof
system is essentially an adaptation of the rely-guarantee reasoning
framework to the setting of weakly isolated database transactions. It
deals with the additional complexity introduced by this setting, while
also taking advantage of the opportunities it provides to simplify the
reasoning. The primary challenge is the weak isolation; how do we
relate a transaction's isolation specification ($\I$) to its rely
relation ($R$) that describes the transaction's environment, so that
the interference is considered only insofar as the isolation level
allows it? Our proof system shows how to compute a rely relation ($R$)
modulo the isolation specification ($\I$) of a transaction. Another
characteristic of the transaction setting is atomicity; a
transaction's writes are not visible to its peers until it commits. In
the context of rely-guarantee, this means that the transactions
guarantee ($G$) should capture the aggregate effect of a transaction,
and not its individual writes.  While shared memory \C{atomic} blocks
also have the same characteristic, the fact that a transactions are
weakly isolated introduces complexity.  Unlike an \C{atomic} block,
the effect of a transaction is \emph{not} a sequential composition of
the effects of its statements because each statement witnesses a
potentially different version of the state. For example,  a series of
operations on a shared variable (\C{X}) can be composed to describe
the effect of an \C{atomic} block, while same cannot be done in a
transaction. Thus, weak isolation needs to be taken into account when
defining and verifying guarantees, and our proof system shows how.
